Planting Predictors of Harvest Yield   by Andrea Brown
========================================================
```{r global_options, include=FALSE}
#Set global options
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r packages}
# Load all of the packages that you end up using in your analysis.

#load the function libraries
library(sp)
library(raster)
library(RColorBrewer)

#loading the plotting libraries
library(ggplot2)
library(dplyr)
library(GGally)
library(gridExtra)
```

# Introduction 
This dataset contains geospatial agricultural data collected during the planting of crops and then again during the harvesting of those planted crops. The planting data sample size is 6,314 and includes the latitude and longitude of where the crop was planted, the seed variety, seeding rate (seeds planted per acre), seed spacing (inches), and speed. The harvest data sample size is 16,626 and includes the latitude and longitude of where the crops were harvested and the harvest yield at that location.

The goal of the following analysis is to match the planting data to the harvest data according to proximity in order to determine the planting variables associated with each harvest yield location. The combined planting and harvest yield data will be used to determine what planting variables, if any, demonstrate correlation to the harvest yield.

#Method
At first I pursued the dataset by trying to calculate the distance between all the points using the "great circle distance," which accounts for the curvature of the earth. However, I quickly realized that there were way too many data points to come up with a solution in a timely manner.

To reduce the amount of computation time, I decided to use the raster library. This is essentially a grid system that stores a value within each grid cell. This reduces the computation time from hours to seconds.

### Raster Creation
In order to create my raster, I write a function using the raster and sp libraries. The function accepts a planting csv file name and harvest csv file name as inputs and returns the combined data frame as the output.

To develop the raster grid for the agricultural field, I convert both the harvest and planting data frames to spatial points data frames using the "sp" library. Then I set the projection to WGS 1984 for the latitude and longitude coordinates. Next I use the raster package to create a grid, or raster, from the spatial extents of the harvest spatial points data frame. I split the grid into 30x30 cells.

Now that I have my raster grid, I use the rasterize function in the "raster" library to convert the planting spatial points data frame to a raster object. Next, I use the raster object I developed to extract the planting variable values at each harvest yield data location. I output a plot of each planting data raster and the combined data frame. It should also be noted that categorical values cannot be converted to a raster, so I transform the seed varietys to an integer during the raster analysis.

```{r extract_crop_yield_function}
extract_crop_yield <- function(planting_filename, 
                               harvest_filename) {
  # Load the Data
  harvest <- read.csv(harvest_filename)
  planting <- read.csv(planting_filename)
  
  # Transform variety from factor levels to integers for raster value extraction
  varieties <- levels(planting$variety)
  planting$variety <- as.integer(planting$variety)
  
  # Convert planting data to spatial points 
  # data frame and set projection
  planting_pts <- planting
  coordinates(planting_pts)=c("long","lat")
  proj4string(planting_pts) <- CRS("+init=epsg:4326")
  
  # Convert coordinates to spatial points 
  # data frame and set projection
  harvest_pts <- harvest
  coordinates(harvest_pts)=c("long","lat")
  proj4string(harvest_pts) <- CRS("+init=epsg:4326")
  
  # Create raster from data extents
  r <- raster(harvest_pts,nrows=30, ncols=30)
  
  # Add values to raster
  planting_raster <- rasterize(planting_pts,r,
                               field=c("variety",
                                       "seeding_rate",
                                       "seed_spacing",
                                       "speed"))
  
  # Plot the planting raster values
  plot(planting_raster$variety, 
     xlab='Longitude',
     ylab='Latitude') +
  title("Seed Variety")
  
  plot(planting_raster$seeding_rate, 
     xlab='Longitude',
     ylab='Latitude') +
  title("Seeding Rate")
  
  plot(planting_raster$seed_spacing, 
     xlab='Longitude',
     ylab='Latitude') +
  title("Seed Spacing")
  
  plot(planting_raster$speed, 
     xlab='Longitude',
     ylab='Latitude') +
  title("Speed")
  
  #extract planting data using raster cell values
  extract_data <- extract(planting_raster,
                    harvest_pts,
                    df=TRUE,
                    sp=TRUE)
  
  #convert spatial data frame to normal data frame
  extract_data <- as.data.frame(extract_data)
  
  # Convert planting variety back to factor levels
  extract_data$variety <- factor(extract_data$variety,labels = varieties)

  return(extract_data)
}
```

### Running the Function
Now I'm ready to run the function and save the new data frame to a csv file called "harvest_planting_data.csv"

```{r run_function_on_csv_files}
data <- extract_crop_yield('planting_sample_data.csv', 'harvest_sample_data.csv')
write.csv(data, file = "harvest_planting_data.csv")
```

### Cleaning the Data
After looking at the raster plots of the extracted planting data, it looks like a few outliers are having a big impact on the data. I remove the top 1% and bottom 1% of values and get rid of one NA value prior to analyzing the dataset.
```{r remove_outliers}
clean_data <- subset(data, !is.na(variety) 
                     & !is.na(seeding_rate) 
                     & !is.na(seed_spacing) 
                     & !is.na(speed))
clean_data <- subset(clean_data,speed > quantile(speed,probs = 0.01)
                     & speed < quantile(speed,probs = 0.99)
                     & seed_spacing > quantile(seed_spacing,probs= 0.01)
                     & seed_spacing < quantile(seed_spacing,probs = 0.99)
                     & seeding_rate > quantile(seeding_rate,probs =0.01)
                     & seeding_rate < quantile(seeding_rate,probs = 0.99))
```

# Exploratory Data Analysis
Now that there is planting-related data at each harvest location, I can begin the exploratory data analysis to identify potential predictors of better harvest yield. Since there is only one dependent variable (yield) and four potential independent variables (variety, seeding rate, seed spacing, and speed), I'll take the time to look in depth at each variable.

```{r distribution_plotting_function}
box_and_hist <- function(data, variable, binwidth, variable_name) {
  grid.arrange(
  ggplot(data = data) +
    geom_histogram(aes(x = variable), binwidth = binwidth, color = "#099DD9") +
    xlab(variable_name) +
    ggtitle('Histogram'),
  
  ggplot(aes(x=1, y=variable), data = data) +
    geom_boxplot( ) +
    ylab(variable_name) +
    ggtitle('Box Plot')
, nrow =1)
}
```

### Yield
I start with the dependent variable: yield. I look at a box plot and histogram of this variable to identify the distribution and any potential outliers. After plotting, although the data is slightly skewed, it appears normally distributed with no extreme outliers.

```{r yield_distribution}
box_and_hist(clean_data, clean_data$yield, 20, 'Crop Yield')
```

### Seed Variety
Planting variety is the only categorical variable in the dataframe. I'll take a look at a box plot and histogram of this variable to identify the distribution and any potential outliers. After plotting, although the data is slightly skewed, it appears normally distributed.

```{r variety_distribution}
grid.arrange(
  ggplot(aes(x=variety, y = yield), data = clean_data) +
    geom_boxplot( ) +
    ggtitle('Seed Variety Box Plot')
, nrow =1)
```

### Seeding Rate
Seeding rate is next up. I'll take a look at a box plot and histogram of this variable to identify the distribution and any potential outliers. After plotting, the data appears to have an outlier of seeding rate = 31.3. The data looks fairly normal without the outliar.

```{r seeding_rate_distribution}
box_and_hist(clean_data, clean_data$seeding_rate, 0.5, 'Seeding Rate')
```

### Seed Spacing
I'll look at seed spacing next. I'll take a look at a box plot and histogram of this variable to identify the distribution and any potential outliers. After plotting, the data appears to have an outlier of seed spacing ~ 6.7 and looks normally distributed if the outlier is removed.

```{r seed_spacing_distribution}
box_and_hist(clean_data, clean_data$seed_spacing, 0.1, 'Seed Spacing')
```

### Planting Speed
Next, I'll look at the speed. I'll take a look at a box plot and histogram of this variable to identify the distribution and any potential outliers. After plotting, the data appears to have a left skew and seems to have a couple outliers of speed < 1. This data looks like it has an exponential distribution.

```{r seed_spacing_distribution}
box_and_hist(clean_data, clean_data$speed, 0.2, 'Speed')
```

# Overview of Bivariate Relationships
I'll create a data frame that does not include the identified outliers. And I'll look at the scatterplot of each variable against the crop yield. I'll also apply an exponential transformation to the axis for the planting speed.

```{r scatterplot_function}
scatter <- function(data, variable, variable_name) {
  ggplot(aes(x = variable, y = yield), data = data) + 
  geom_point(size = 1, alpha = 0.5, position = 'jitter') +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  labs(x = variable_name, y = "Crop Yield")
}
```

### Seed Variety
I'll look at a box plot again and add the mean. It looks like seed variety "DKC63-33RIB" shows slightly higher crop yield than seed variety "P1498."

```{r seed_variety_box_plot_and_mean}
grid.arrange(
  ggplot(aes(x=variety, y = yield), data = clean_data) +
    geom_boxplot( ) +
    stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
    ggtitle('Seed Variety Box Plot')
, nrow =1)
```

### Seeding Rate
The seeding rate does not look very correlated to crop yield.
```{r seeding_rate_scatterplot}
scatter(clean_data, clean_data$seeding_rate,"Seeding Rate")
cor.test(clean_data$seeding_rate, clean_data$yield)
```

### Seed Spacing
It looks like closer seed spacing may be correlated to higher crop yield.
```{r seeding_rate_scatterplot}
scatter(clean_data, clean_data$seed_spacing,"Seed Spacing")
cor.test(clean_data$seed_spacing, clean_data$yield)
```

### Planting Speed
Planting speed does not appear very correlated to the crop yield.
```{r seeding_rate_scatterplot}
scatter(clean_data, clean_data$speed,"Speed")
cor.test(clean_data$speed, clean_data$yield)
```

# Multi-variate Plots
Next, I'll take a closer look at my two most promising variables: seed spacing and variety.

```{r multivariate_plot}
ggplot(aes(x = seed_spacing, y = yield, color = variety), data = clean_data) + 
  geom_point(size = 1, position = 'jitter') +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  scale_color_brewer(type = "qual", palette = 1,
    guide = guide_legend(title="Seed Variety",
    override.aes = list(alpha = 0.5, size = 2))) +
  labs(x = "Seed Spacing", y = "Crop Yield") +
  ggtitle('Correlation between Seed Spacing, Seed Variety, and Crop Yield')
```

#Regression Model
It's fairly difficult to observe any trends because There's so few points with seed variety "P1498" and not much variation in seed spacing for that seed variety. I'll try to create a regression model out of these variables.
```{r model_1}
m1 <- lm(yield ~ seed_spacing, data = clean_data)
summary(m1)
```

```{r model_2}
m2 <- lm(yield ~ I(seed_spacing) + I(as.integer(variety)), data = clean_data)
summary(m2)
```

```{r model_3}
m3 <- lm(yield ~ I(as.integer(variety)), data = clean_data)
summary(m3)
```

#Analysis of Model Error
Plugging in the resulting linear regression equation and plotting the error reveals a normally distributed data set. But the error is fairly high compaired to the yield values.
```{r model_prediction}
clean_data$pred_yield <- 245.1841 - 12.9005*as.integer(clean_data$variety)
```

```{r model_error}
clean_data$SE <- (clean_data$yield - clean_data$pred_yield)
ggplot(data = clean_data) + 
  geom_histogram(aes(x = SE,colour = "red"),show.legend = F, binwidth = 0.1) +
  ggtitle('Error in Predicted Yield')
```

### Conclusions
After extracting the planting data into the locations where crop yield was recorded, I analyzed the correlations between the planting data and the assocationed crop yield data. After observing the distributions of the variables and removing outliers, it appeared that seed spacing and seed variety may exhibit correlation to crop yield. Insertion of each variable into a linear regression model demonstrated that reduced seed spacing may be correlated to higher crop yields. Seed variety was not very helpful in predicting the crop yield, which may be due to the small number of data points for seed variety P1498. While the error of the linear regression model is normal, the standard error and the model F-statistic and R-squared values indicate the model is not very good.

### Future Work
This model has a lot of room for improvement. Ideas to improve the model as well as improve the identification of variables correlated to higher crop yields include:
- Instead of extracting the planting data into the harvest yield, extract the harvest yield into the planting data. There will be a much smaller sample size, but the yield data may more precisely be associated with the planting data.
- Decrease the cell size of the raster before extracting the planting data.
- Collect additional data with the harvest yield, such as the soil moisture, nitrate, phosphorus, etc.
- Collect additional data throughout the crop cycle. For example, collect monthly soil moisture data and soil temperature data.